Outlier Detection and Comparison of Origin-Destination Flows using Data Depth

### 2018-05-30 GIScience_2018/giscience_r2/giscience/giscience2018.tex
cc

\begin{abstract}
Advances in location-aware technology have resulted in the generation of a huge volume of trajectory data.
Origin-destination (OD) trajectories provide rich information on urban flow and transport demand.
This study presents a new methodology to detect OD flow outliers and conduct hypothesis testing between two OD flows in terms of the variations of spatial extent, namely, spread.
The proposed method is based on data depth, which measures the centrality and outlyingness of a point with respect to a given dataset in $\mathbb{R}^d$.
Based on the center-outward ordering property, the proposed method analyzes the underlying characteristics of OD flows, such as location, outlyingness, and spread.
The ability of the proposed method to detect OD anomalies is compared with that of the Mahalanobis distance approach, and an F-test is used to verify the difference in scale. Empirical evaluation has demonstrated that the proposed method effectively identifies OD flows outliers in an interactive way. Furthermore, the proposed method can provide new perspectives by considering the overall structure of data when comparing two different OD flows in terms of scale.

@@

As an alternative method, the state-of-the-art Mahalanobis distance approach detected similar outlier OD flows.
However, the number of outliers detected was different.
This occurred because our OD flows data had heavy tail distributions, that is, many of the OD flows with a long distance from the depth median depicted in Figure \ref{fig:weekdays_rare}. Mahalanobis distance is known to be inadequate when the underlying data have heavy tail distributions \cite{wilcox12Book}. Thus, the presence of outliers may mask the detection of other outliers in Mahalanobis distance approach. Furthermore, it can only detect OD flow outliers with high numbers of trips during time $t_1$ and time $t_2$. It is difficult to detect OD flows outliers that have different properties, such as substantial differences in the number of trips when comparing with time $t_1$ and time $t_2$.

@@

This paper provides a new methodology for identifying outlier OD flows and the difference in scale between two different OD flows at $t_1$ and $t_2$.
The proposed method is based on the concept of data depth.
Data depth is robust statistics, which is suited to non-Gaussian distribution of the underlying datasets.
Compared with standard statistics, this approach enhances understanding of the differences and the magnitude of the differences between tow OD flows.

@@

This study made no attempt to consider geographic contexts such as locational circumstances or surrounding environment in understanding OD flows. 
Further investigation will focus on integrating the analysis of OD flows with geographic context.
Such an effort will lead to knowledge discovery and understanding the dynamics of urban flow.

>>>>

\begin{abstract}
Advances in location-aware technology have lead to the generation of a huge volume of trajectory data.
Origin-destination (OD) trajectories provide rich information on urban flow and transport demand.
This study presents a new methodology to detect OD flow outliers and conduct hypothesis testing between two OD flows in terms of the variations of spatial extent, namely, spread.
The proposed method is based on data depth, which measures the centrality and outlyingness of a point with respect to a given dataset in $\mathbb{R}^d$.
Based on the center-outward ordering property, the proposed method analyzes the underlying characteristics of OD flows, such as location, outlyingness, and spread.
The ability of the proposed method to detect OD anomalies is compared with that of the Mahalanobis distance approach, and an F-test is used to verify the difference in scale. Empirical evaluation has demonstrated that the proposed method effectively identifies OD flows outliers in an interactive way. Furthermore, the proposed method can provide new perspectives by considering the overall structure of data when comparing two different OD flows in terms of scale.

@@

As an alternative method, the state-of-the-art Mahalanobis distance-based approach detected similar outlier OD flows.
However, the number of outliers detected was different.
This occurred because our OD flows data had heavy tail distributions, that is, many of the OD flows with a long distance from the depth median depicted in Figure \ref{fig:weekdays_rare}. 
Mahalanobis distance is known to be inadequate when the underlying data have heavy tailed distributions \cite{wilcox12Book}. 
Thus, the presence of outliers may mask the detection of other outliers in Mahalanobis distance approach.
Furthermore, it can only detect OD flow outliers with high numbers of trips during time $t_1$ and time $t_2$. 
It is difficult to detect OD flows outliers that have different properties, such as substantial differences in the number of trips when comparing with time $t_1$ and time $t_2$.

@@

This paper provides a new methodology for identifying outlier OD flows and the difference in scale between two different OD flows at $t_1$ and $t_2$.
The proposed method is based on the concept of data depth.
Data depth is robust statistics, which is suited to non-Gaussian distribution of the underlying datasets.
Compared with standard statistics, this approach enhances understanding of the differences and the magnitude of the differences between two OD flows.

@@

This study made no attempt to incorporate geographic contexts such as locational circumstances or surrounding environment in understanding OD flows. 
Further investigations will focus on integrating the analysis of OD flows with geographic context.
Such an effort will lead to knowledge discovery and better understandings regarding the dynamics of urban flow.
